---
title: "MK772 — 3‑Compartment Transit PK: Reconstruction & Design"
format:
  revealjs:
    slide-number: true
    code-overflow: wrap
    code-line-numbers: true
    incremental: true
    footer: "nlmixr2 • rxode2 • Quarto"
execute:
  echo: true
  warning: false
  message: false
  cache: true
editor: source
---

# 0. Setup

```{r}
#| label: setup
#| include: false
# Packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(readr)
  library(glue)
  library(nlmixr2)
  library(nlmixr2est)
  library(broom.mixed)
  library(gt)
  library(stringr)
  library(patchwork)
  library(xpose)
  library(xpose.nlmixr2)
})

# ---- Plot theme & defaults ----
# A clean, presentation-friendly theme and sensible defaults

theme_mk772 <- function(base_size = 14, base_family = "") {
  ggplot2::theme_minimal(base_size = base_size, base_family = base_family) +
    ggplot2::theme(
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_line(linewidth = 0.3),
      strip.background = ggplot2::element_rect(fill = "#f6f7f9", color = NA),
      strip.text = ggplot2::element_text(face = "bold"),
      plot.title.position = "plot",
      plot.title = ggplot2::element_text(face = "bold"),
      legend.title = ggplot2::element_text(),
      legend.position = "right"
    )
}

# Set the theme globally for the entire notebook/presentation
ggplot2::theme_set(theme_mk772())

# Sensible defaults for geoms (can tweak later)
try(ggplot2::update_geom_defaults("line",  list(linewidth = 0.7, alpha = 0.9)), silent = TRUE)
try(ggplot2::update_geom_defaults("point", list(size = 1.2,  alpha = 0.8)), silent = TRUE)

# Helper scale used a lot in PK plots
scale_y_conc <- function(...) ggplot2::scale_y_continuous(trans = scales::pseudo_log_trans(), ...)

# Helper: load first existing file
first_existing <- function(paths) paths[file.exists(paths)][1]

# Standardize columns for plotting/fit
standardize_pk <- function(df){
  # Try to discover the DV column (concentration)
  dv_col <- dplyr::case_when(
    "DV" %in% names(df) ~ "DV",
    "DV_CP" %in% names(df) ~ "DV_CP",
    "Cp" %in% names(df) ~ "Cp",
    TRUE ~ NA_character_
  )
  stopifnot(!is.na(dv_col))
  df %>%
    rename(
      ID   = tidyselect::any_of(c("ID","id")),
      TIME = tidyselect::any_of(c("TIME","time"))
    ) %>%
    mutate(DV = .data[[dv_col]]) %>%
    arrange(ID, TIME)
}

# Paths (relative to repo root)
data_paths <- c(
  "../data/pkpd_3cmt_transit.rds",
  "../data/pkpd_3cmt_transit.csv"
)

fit <- read_rds("../results/fit_3cmt_transit.rds")

path_data <- first_existing(data_paths)
if (is.na(path_data)) stop("No data file found in expected locations.")

raw <- if (grepl("\\.rds$", path_data)) readr::read_rds(path_data) else readr::read_csv(path_data, show_col_types = FALSE)

pk <- standardize_pk(raw)

# A lightweight nlmixr2 model definition matching the 3‑compartment + transit structure
pk3cmt_transit <- function(){
  ini({
    tcl   <- log(14)       # CL (L/h) baseline
    tvc   <- log(60)       # Vc (L)
    tktr  <- log((3+1)/10) # ktr derived from NTR=3, MTT=10h (modify during refinement)

    eta.cl ~ 0.09
    eta.vc ~ 0.04

    prop.err.cp <- 0.07
  })
  model({
    # Fixed peripherals; keep simple for demo; you can estimate later
    Q2   <- 6.0
    Q3   <- 5.0
    Vp1  <- 30
    Vp2  <- 40

    # Covariate: allometry on CL (expects WT in dataset; if absent, assumes 70 kg)
    WT1 <- ifelse(exists("WT"), WT, 70)
    CL <- exp(tcl + eta.cl) * (WT1/70)^0.75
    Vc <- exp(tvc + eta.vc)
    ktr <- exp(tktr)

    # Compartments: A0, TR1-TR3, A1 (central), A2/A3 (peripherals)
    d/dt(A0)  = -ktr * A0;
    d/dt(TR1) =  ktr * A0 - ktr * TR1;
    d/dt(TR2) =  ktr * TR1 - ktr * TR2;
    d/dt(TR3) =  ktr * TR2 - ktr * TR3;

    d/dt(A1) = ktr * TR3 - (CL / Vc) * A1
               - Q2 / Vc * A1 + Q2 / Vp1 * A2
               - Q3 / Vc * A1 + Q3 / Vp2 * A3;
    d/dt(A2) = Q2 / Vc * A1 - Q2 / Vp1 * A2;
    d/dt(A3) = Q3 / Vc * A1 - Q3 / Vp2 * A3;

    Cp <- A1 / Vc
    Cp ~ prop(prop.err.cp) | A1
  })
}
```

---

# 0.1 Data overview

## What’s loaded

```{r}
#| label: data-overview
#| tbl-cap: "Head of standardized PK data"
cat(glue("Using data file: {path_data}\n"))

pk %>% 
  mutate(across(c(ID), as.integer)) %>%
  head(10)
```

## Quick counts

```{r}
#| label: data-counts
pk_counts <- pk %>% summarise(n_subjects = n_distinct(ID), n_obs = sum(is.finite(DV)))
pk_counts
```

## Spaghetti plot (random sample of subjects)

```{r}
#| label: spaghetti
set.seed(1)
ids <- pk %>% distinct(ID) %>% slice_sample(n = 12) %>% pull(ID)

pk %>%
  filter(ID %in% ids) %>%
  ggplot(aes(TIME, DV, group = ID)) +
  geom_line(alpha = 0.6) +
  geom_point(size = 0.6, alpha = 0.6) +
  facet_wrap(~ID, scales = "free_y") +
  scale_y_continuous(trans = scales::pseudo_log_trans()) +
  labs(title = "Observed concentrations (subset)", x = "Time", y = "DV (concentration)") +
  theme_bw(base_size = 12)
```

## Sampling time distribution

```{r}
#| label: time-hist
pk %>%
  filter(is.finite(TIME)) %>%
  ggplot(aes(TIME)) +
  geom_histogram(bins = 40) +
  labs(title = "Sampling time histogram", x = "Time", y = "Count") +
  theme_bw(base_size = 12)
```

---

# 0.2 Demonstrate model fit (estimates in tidy format)

> Goal here is to *show* a fit and a clean parameter table. You can load a saved fit if you have one, or run SAEM inline (may take a few minutes). On slides, prefer loading a saved result.

## Traceplots 

```{r}
#| label: saem-traceplots
#| message: false
#| warning: false
#| fig-cap: "SAEM parameter traceplots"

library(nlmixr2)      # re-exports traceplot()
library(nlmixr2plot)  # plotting backend

traceplot(fit) + labs(x = "Iteration", y = "Value") + theme_mk772()
```


## Tidy parameter table

```{r}
#| label: tidy-pars
#| tbl-cap: "Fixed effects (back-transformed where applicable) and variance components"
library(broom.mixed)
library(gt)

# read in all the values for the paramters 
fx <- tidy(fit, effects = "fixed")
rp <- tidy(fit, effects = "ran_pars") |> dplyr::filter(group=="ID")  
sg <- tidy(fit, effects = "ran_pars") |> dplyr::filter(group!="ID")  

# # --- Fixed effects: back-transform log-params ---
bt_map <- c(tcl = "CL", tvc = "Vc", tktr = "ktr")
fixed_tbl <- fx %>%
  transmute(
    term,
    estimate,
    se = dplyr::coalesce(std.error, NA_real_)
  ) %>%
  mutate(
    Parameter   = dplyr::recode(term, !!!bt_map, .default = term),
    Estimate_bt = if_else(term %in% names(bt_map), exp(estimate), estimate),
    SE_bt       = if_else(term %in% names(bt_map) & !is.na(se), exp(estimate) * se, se),
    LCI         = if_else(!is.na(SE_bt), Estimate_bt * exp(-1.96 * dplyr::coalesce(se, 0)), NA_real_),
    UCI         = if_else(!is.na(SE_bt), Estimate_bt * exp( 1.96 * dplyr::coalesce(se, 0)), NA_real_)
  ) %>%
  transmute(
    Component = "Fixed effects",
    Parameter,
    Estimate  = Estimate_bt,
    SE        = SE_bt,
    LCI, UCI,
    Notes     = NA_character_
  )

# --- Random effects: parse SDs (diag) and correlations (off-diag) ---
# broom.mixed typically returns columns like: effect, group, term, estimate, component (\"sd\", \"cor\", or \"sdcor\")
iiv_var <- NULL
iiv_cor <- NULL


# SDs -> convert to variances; add %CV note (SD ~ omega on ETA-scale)
iiv_var <- rp |> 
    transmute(
        Component = "Random effects (Ω)",
        Parameter = dplyr::case_when(
            term == "sd__eta.cl" ~ "IIV CL",
            term == "sd__eta.vc" ~ "IIV Vc",
            TRUE ~ term),
        Estimate  = estimate^2,  # variance
        SE        = NA_real_, LCI = NA_real_, UCI = NA_real_,
        Notes     = paste0("CV%≈ ", sprintf("%.1f", estimate * 100))
        )

# --- Residual (Σ) ---
sigma_tbl <- sg |> 
    transmute(
      Component = "Residual error (Σ)",
      Parameter = "Proportional",
      Estimate  = estimate,
      SE = NA_real_,
      Notes = NA_character_
    )

# --- Combine and render ---
all_params <- bind_rows(fixed_tbl, iiv_var, iiv_cor, sigma_tbl) %>%
  mutate(Component = factor(Component, levels = c("Fixed effects","Random effects (Ω)","Residual error (Σ)"))) %>%
  arrange(Component, desc(str_detect(Parameter, "^IIV\\(")), Parameter)

all_params %>%
  arrange(factor(Component, levels = c("Fixed effects","Random effects (Ω)","Residual error (Σ)")),
          desc(grepl("^IIV\\(", Parameter)),  # optional ordering
          Parameter) %>%
  gt(groupname_col = "Component") %>%   # <-- turns Component into group headers
  cols_hide("Component") %>%            # <-- hides the repeating column
  fmt_number(c(Estimate, SE, LCI, UCI), decimals = 3) %>%
  tab_options(row_group.as_column = FALSE) %>%  # default; keeps headers in the margin
  opt_row_striping()

```

## Goodness of fit 1 -- Predicted v/s DV

```{r}
#| label: pred v/s DV
#| eval: true
#| message: false


# prepare data for the gof plots
library(xpose)
library(xpose.nlmixr2)
library(patchwork)

# load the gof data and xpose it
xp_fit <- xpose_data_nlmixr2(fit)

plt1 <- (
    dv_vs_pred(xp_fit, type = "p") + 
        labs(title = "", subtitle = "", caption = "") + 
        theme_mk772()
        )
plt2 <- (
    dv_vs_ipred(xp_fit, type = "p") +
        labs(title = "", subtitle = "", caption = "") + 
        theme_mk772()
    )

plt1 | plt2
```

## Goodness of fit 2 -- IWRES

```{r}

#| label: pred v/s DV
#| eval: false
#| message: false
plt1 <- (
    res_vs_pred(xp_fit, res = "IWRES", type="p") + 
        labs(title = "", subtitle = "", caption = "") + 
        theme_mk772()
        )


xp_tad <- set_var_types(xp_fit, idv = "TAD") 
        
plt2  <- (
    res_vs_idv(xp_tad, res = "IWRES", type = "p") + 
        labs(title = "", subtitle = "", caption= "") + 
        theme_mk772()
        )


plt1 | plt2
```
---

## Goodness of fit 3 -- iwres

```{r}
#| label: iwres-health
#| message: false
#| warning: false

library(xpose)

p_iwres_hist <- res_distrib(
  xp_fit, res = "IWRES", type = "h", caption = NULL) +
    labs(y = "Density", title = "", subtitle = "", caption = "") +
    theme_mk772() 

p_iwres_qq <- res_qq(
  xp_fit, res = "IWRES", type = "p", caption = NULL) + 
    labs(title = "", subtitle = "", y = "Quantiles of IWRES") +
    theme_mk772() 

(p_iwres_hist | p_iwres_qq) 
```

## Goodness of fit 4 -- eta 

```{r}
#| label: eta-health
#| message: false
#| warning: false

library(xpose.nlmixr2)
library(xpose)
library(patchwork)

p_eta_hist <- (
    eta_distrib(
        xp_fit, type = "h", ncol = 2, 
        nrow = 1, page = 1) +
    labs(title = "", subtitle = "", caption = "", 
         x = "Value", y = "Density") + 
    theme_mk772()
    )

p_eta_qq <- (
    eta_qq(
        xp_fit, type = "p", ncol = 2, 
        nrow = 1, page = 1, caption = NULL) + 
    labs(title = "", subtitle = "") + 
    theme_mk772()
    )

# 2×2 layout: top row = two ETA histograms; bottom row = two ETA QQ plots
(p_eta_hist / p_eta_qq)
```

## VPC 1: Visual predictive checks

```{r}
#| label: vpc-pc
#| message: false
#| warning: false
#| fig-cap: "Prediction-corrected VPC (n = 300)"

library(nlmixr2plot)
library(patchwork)

set.seed(42)

vpc1 <- vpcPlot(
  fit,                 # your nlmixr2 SAEM fit object
  n = 300,             # number of simulation replicates
  pred_corr = TRUE,    # prediction-corrected VPC
  idv = "time",         # independent variable column in your data
  show=list(obs_dv=TRUE)
) + labs(x = "Time", y = "DV") + theme_mk772()

vpc2 <- vpcPlot(
  fit,                 # your nlmixr2 SAEM fit object
  n = 300,             # number of simulation replicates
  pred_corr = TRUE,    # prediction-corrected VPC
  idv = "tad",         # independent variable column in your data
  show=list(obs_dv=TRUE)
) + labs(x = "Time after dose", y = "DV") + theme_mk772()


vpc1 / vpc2
```
 
## VPC 2: Individual fit:

```{r}
 
 df <- fit |> as_tibble()

  # for reproducibility
ids_10 <- df |> distinct(ID) |> slice_sample(n = 6) |> pull(ID)
ind_df <- df |> filter(ID %in% ids_10)

plt <- ind_df |>
  ggplot(aes(x = TIME)) +
  geom_point(aes(y = DV), alpha = 0.6, size = 1) +
  geom_line(aes(y = IPRED, linetype = "IPRED")) +
  geom_line(aes(y = PRED,  linetype = "PRED")) +
  scale_linetype_manual(values = c(IPRED = 1, PRED = 2), name = "") +
  facet_wrap(~ ID, ncol = 2, scales = "free_y") +
  labs(x = "Time", y = "DV", title = "Individual fits (6 random IDs)")+
  theme_mk772() +
  theme(legend.position = "bottom")

plt

```  


# Next steps (placeholders for new chats)


- **2) Uncertainty** — bootstrap for parameters and exposures. -- Not doing it
- **3) Regimen simulations** — exposure tables (AUC, Cmax, Ctrough).
- **4) PTA curves** — probability of target attainment vs regimen.
- **5) Power/sample size** — simulation-based operating characteristics.
- **6) Optimal sampling** — schedule tweaks and RSE impact.
- **7) Robustness** — sensitivity to NTR/MTT and error model.
- **8) Shiny** — wrap dose/interval/WT → exposure/PTA UI.
